Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-20 paper/paper.tex
Finished the text; figures next.

\begin{abstract}
Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes. Path indexes are one of the most important tools for working with variation graphs. They generalize text indexes for graphs, allowing one to find the paths matching the query string. We propose using pruned de Bruijn graphs as path indexes, encoding them space-efficiently with the Burrows-Wheeler transform. We also generalize many ideas from text indexing literature to work with graphs. The proposed approach has been implemented in the variation graph toolkit vg.
\end{abstract}

@@

In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

We implement query $G'.\find(X)$ with \emph{backward searching}. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes of the path graph. Now assume that $\abs{X} \ge 1$ and that query $G'.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range corresponding to the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i}], X[i])$.

@@

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction. The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index. There are two main reasons: (1) we use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}; and (2) $\locate$ queries generate duplicates, which must be filtered out. Switching to the simplified encoding for maximally pruned graphs would make $\find$ queries faster and $\locate$ queries slower.

@@

The compressed suffix tree used in GCSA2 may be useful for indexing texts as well, as it is both simple to implement and fast in practice. As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently. We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.

% FIXME from here
% additional edges using hypertext index; haplotypes and potential recombinations

% graph formalism, grammars

>>>>

\begin{abstract}
Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes. Path indexes are one of the most important tools for working with variation graphs. They generalize text indexes for graphs, allowing one to find the paths matching the query string. We propose using pruned de Bruijn graphs as path indexes, and encode them space-efficiently with the Burrows-Wheeler transform. We also generalize many ideas from text indexing literature to work with graphs. The proposed approach has been implemented in the variation graph toolkit vg.
\end{abstract}

@@

Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

We implement query $G'.\find(X)$ with \emph{backward searching}. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes of the path graph. Now assume that $\abs{X} \ge 1$ and that query $G'.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range corresponding to the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i}], X[i])$.

@@

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned before index construction. The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index. There are two main reasons: (1) we use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}; and (2) $\locate$ queries generate duplicates, which must be filtered out. Switching to the simplified encoding for maximally pruned graphs would make $\find$ queries faster and $\locate$ queries slower.

@@

The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice. As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently. We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.

We often have to prune complex regions of the input graph before indexing it. This causes false negatives: paths that exist in the input graph but not in the index. We could adapt techniques from the hypertext index \cite{Thachuk2013} to avoid the false negatives. We create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph. Then we create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph. We then search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index. For all $0 < i < \abs{X}$, we can combine the partial matches $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ into complete matches crossing one recombination edge with a range query in the edge matrix.

While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately. For example, if sequence $S$ can occur in different positions of the genome (e.g.~$XSYZ$ and $XYSZ$), we can have a single copy or multiple copies of $S$ in the reference. Neither of the options is good for a reference genome. With a single copy, we create paths in the reference that do not correspond to any valid genome. With the second option, we duplicate sequences that are not only identical but also functionally the same. One solution is to switch from graphs to \emph{context-free grammars}. As long as the grammar is non-nested, we can handle it with the hypertext index. We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
